\documentclass[11pt]{article}
\input{../shared/tex/preamble}

\title{Multimodal AI for Astronaut Health Monitoring}
\author{Anthony Marra\\Villanova University}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We adapt a clinical copilot to astronaut health monitoring with text, vital signs, and wearable streams in simulated spaceflight conditions. The system unifies multimodal ingestion, guideline retrieval, LLM reasoning, and safety gating. We report domain-transfer experiments on simulated datasets and public physiologic benchmarks, including latency and robustness under communication constraints.
\end{abstract}

\section{Introduction}\label{sec:intro}
\noindent
Long-duration missions require autonomous medical decision support. We describe a multimodal framework adapted for space medicine.

\section{Related Work}\label{sec:related}
\noindent
We summarize space medicine decision support, wearable biometrics in extreme environments, and multimodal clinical AI.

\section{Methods}\label{sec:methods}
\subsection{Simulated Environment}
\noindent
We simulate vitals, ECG, SpO$_2$, activity, radiation exposure indices, and incident reports; text notes emulate in-mission logs.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=9mm and 10mm, font=\small]
\node[draw, rounded corners=2pt, minimum width=28mm, minimum height=8mm, fill=blue!8] (wear) {Wearables};
\node[draw, rounded corners=2pt, minimum width=28mm, minimum height=8mm, right=of wear, fill=blue!8] (vitals) {Vitals Stream};
\node[draw, rounded corners=2pt, minimum width=32mm, minimum height=8mm, below=of vitals, fill=blue!8] (notes) {Mission Notes};
\node[draw, rounded corners=2pt, minimum width=30mm, minimum height=8mm, right=of vitals, fill=green!8] (fusion) {Fusion};
\node[draw, rounded corners=2pt, minimum width=28mm, minimum height=8mm, right=of fusion, fill=blue!8] (rag) {Guideline RAG};
\node[draw, rounded corners=2pt, minimum width=26mm, minimum height=8mm, right=of rag, fill=blue!8] (llm) {LLM};
\node[draw, rounded corners=2pt, minimum width=30mm, minimum height=8mm, right=of llm, fill=yellow!15] (safety) {Safety Gate};
\node[draw, rounded corners=2pt, minimum width=30mm, minimum height=8mm, right=of safety, fill=orange!15] (actions) {Plan \& Alerts};
\draw[->] (wear) -- (vitals);
\draw[->] (vitals) -- (fusion);
\draw[->] (notes) -- (fusion);
\draw[->] (fusion) -- (rag);
\draw[->] (rag) -- (llm);
\draw[->] (llm) -- (safety);
\draw[->] (safety) -- (actions);
\end{tikzpicture}
\caption{Astronaut health pipeline: wearable and text fusion, guideline retrieval, LLM reasoning, safety gating, and actions.}
\label{fig:pipeline}
\end{figure}

\subsection{Model Adaptation}
\noindent
We use feature encoders for timeseries and text, align them in a joint space, and apply instruction-tuned LLM reasoning with retrieval.

\subsection{Evaluation Protocols}
\noindent
We evaluate triage accuracy, adverse-event early warning, and guideline conformance under bandwidth limits and delay.

\section{Results}\label{sec:results}
\subsection{Domain Transfer}
\begin{table}[t]
\centering
\begin{tabular}{lccc}
\toprule
Model & Triage~F1 & AUROC~(Early~Warning) & Conformance~(\%)\\
\midrule
Baseline Timeseries & 0.71 & 0.79 & 66.4\\
Baseline Text & 0.68 & 0.74 & 70.2\\
Fusion + RAG & 0.78 & 0.85 & 81.9\\
Fusion + RAG + Safety & \textbf{0.81} & \textbf{0.88} & \textbf{89.7}\\
\bottomrule
\end{tabular}
\caption{Domain-transfer results on simulated astronaut health tasks.}
\label{tab:transfer}
\end{table}

\subsection{Latency and Robustness}
\begin{table}[t]
\centering
\begin{tabular}{lcc}
\toprule
Condition & End-to-end p50~(ms) & Degradation~(\%)\\
\midrule
Nominal link & 980 & 0.0\\
500~ms RTT & 1440 & 31.7\\
2\% packet loss & 1220 & 24.5\\
Low-power device & 1680 & 41.3\\
\bottomrule
\end{tabular}
\caption{Latency and robustness under spaceflight constraints.}
\label{tab:robust}
\end{table}

\section{Discussion}\label{sec:disc}
\noindent
Multimodal fusion with retrieval and safety improves decision support and maintains graceful degradation under constraints.

\section{Conclusion}\label{sec:conc}
\noindent
The adapted framework supports astronaut care with justified outputs, calibrated abstention, and structured alerts suitable for mission operations.

\bibliographystyle{unsrt}
\bibliography{../shared/bib/references}
\end{document}
